\pkg{systemfit}: A Package for Estimating Systems of Simultaneous Equations in \proglang{R}

### 2008-10-16 pkg/inst/doc/systemfit.Rnw
changed encoding to utf-8 and added the line "\usepackage[utf8]{inputenc}"
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@969 591db2bd-e0a0-4d22-8c19-2192dbb7da31
removed "systemfit" from \VignetteDepends{}, because this gives an error message in R 2.6.2 (and possibly older versions of R) but not in R 2.7.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@985 591db2bd-e0a0-4d22-8c19-2192dbb7da31

\documentclass[article,nojss]{jss}
\usepackage{csquotes}
\MakeOuterQuote{°}
\usepackage{amsmath}
\usepackage{float}

@@

% initialisation stuff
\SweaveOpts{engine=R}
%\VignetteIndexEntry{Systemfit}
%\VignetteDepends{systemfit,plm,sem}
%\VignetteKeywords{R, system of simultaneous equations,
% seemingly unrelated regression, two-stage least squares,
% three-stage least squares, instrumental variables}
%\VignettePackage{systemfit}
<<echo=FALSE>>=
options( prompt = "R> ", ctinue = "+  " )
@

@@

Many theoretical models that are econometrically estimated
consist of more than one equation.
The disturbance terms of these equations are likely
to be contemporaneously correlated,
because unconsidered factors
that influence the disturbance term in one equation
probably influence the disturbance terms in other equations, too.
Ignoring this contemporaneous correlation
and estimating these equations separately
leads to inefficient estimates of the coefficients.
However, estimating all equations simultaneously
with a °generalized least squares° (GLS) estimator,
which takes the covariance structure of the residuals into account,
leads to efficient estimates.
This estimation procedure is generally called
°seemingly unrelated regression° \citep[SUR,][]{zellner62}.
Another reason to estimate a system of equations simultaneously are
cross-equation restrictions on the coefficients.%
\footnote{
Especially the economic theory suggests many cross-equation
restrictions on the coefficients (e.g., the symmetry restriction in demand models).
}
Estimating the coefficients under cross-equation restrictions
and testing these restrictions
requires a simultaneous estimation approach.

Furthermore, these models can contain variables
that appear on the left-hand side in one equation
and on the right-hand side of another equation.
Ignoring the endogeneity of these variables can lead to inconsistent
estimates.
This simultaneity bias can be corrected for by
applying a °two-stage least squares° (2SLS) estimation
to each equation.
Combining this estimation method with the SUR method results
in a simultaneous estimation of the system of equations
by the °three-stage least squares° (3SLS) method
\citep{zellner62b}.

@@

The \pkg{systemfit} package provides the capability to estimate
systems of linear equations in \proglang{R}
\citep{r-project-07}.
Currently, the estimation methods
°ordinary least squares° (OLS),
°weighted least squares° (WLS),
°seemingly unrelated regression° (SUR),
°two-stage least squares° (2SLS),
°weighted two-stage least squares° (W2SLS), and
°three-stage least squares° (3SLS)
are implemented.%
\footnote{
In this context,
the term °weighted° in °weighted least squares° (WLS) and
°weighted two-stage least squares° (W2SLS) means
that the \emph{equations} might have different weights
and \emph{not}
that the \emph{observations} have different weights.
}
The WLS, SUR, W2SLS, and 3SLS estimates can be based
either on one-step (OLS or 2SLS) (co)variances
or these estimations can be iterated,
where the (co)variances are calculated from the estimates of the previous step.
Furthermore,
the \pkg{systemfit} package
provides statistical tests for restrictions on the coefficients
and for testing the consistency of the 3SLS estimation.

@@

We can write the °stacked° system as
\begin{equation}
   \left[ \begin{array}{c}
      y_1 \\ y_2\\ \vdots\\ y_G
   \end{array} \right] = 
   \left[ \begin{array}{cccc}
      X_1 & 0 & \cdots & 0\\
      0 & X_2 & \cdots & 0\\
      \vdots & \vdots & \ddots & \vdots\\
      0 & 0 & \cdots & X_G
   \end{array}\right]
   \left[ \begin{array}{c}
      \beta_1 \\ \beta_2 \\ \vdots\\ \beta_G
   \end{array} \right] +
   \left[ \begin{array}{c}
      u_1 \\ u_2 \\ \vdots\\ u_G
   \end{array} \right]
   \label{eq:model-array}
\end{equation}

@@

If the disturbances are contemporaneously correlated, a generalized
least squares (GLS) estimation leads to an efficient estimator for the coefficients.
In this case, the GLS estimator is generally called °seemingly unrelated regression°
(SUR) estimator \citep{zellner62}.
However, the true covariance matrix of the disturbance terms
is generally unknown.
The textbook solution for this problem is
a feasible generalized least squares (FGLS) estimation.
As the FGLS estimator is based on
an estimated covariance matrix of the disturbance terms,
it is only asymptotically efficient.
In case of a SUR estimator, $\OHat$ in Equations~\ref{eq:ols-wls-sur}
and~\ref{eq:cov-ols-wls-sur}
is $\SHat \otimes I_T$,
where $\SHat$ is the estimated covariance matrix of the disturbance terms.

@@

At the first stage new (°fitted°) regressors are obtained by
\begin{equation}
   \XHat_i = Z_i \left( Z_i^\top Z_i \right)^{-1} Z_i^\top X_i .
\end{equation}
Then, these °fitted° regressors are substituted for the original regressors
in Equation~\ref{eq:ols-wls-sur}
to obtain unbiased 2SLS, W2SLS, or 3SLS estimates of $\beta$ by
\begin{equation}
   \bHat
   = \left( \XHat^\top \OHat^{-1} \XHat \right)^{-1} \XHat^\top \OHat^{-1} y ,
   \label{eq:2sls-w2sls-3sls}
\end{equation}
where
\begin{equation}
   \XHat =
   \left[ \begin{array}{cccc}
      \XHat_1 & 0       & \cdots & 0\\
      0       & \XHat_2 & \cdots & 0\\
      \vdots  & \vdots  & \ddots & \vdots\\
      0       & 0       & \cdots & \XHat_G
   \end{array}\right] .
\end{equation}
An estimator of the covariance matrix of the estimated coefficients
can be obtained from Equation~\ref{eq:cov-ols-wls-sur} analogously.
Hence, we get
\begin{equation}
   \COVHat \left[ \bHat \right]
   = \left( \XHat^\top \OHat^{-1} \XHat \right)^{-1} .
   \label{eq:cov-2sls-w2sls-3sls}
\end{equation}

@@

If the disturbances are contemporaneously correlated,
a feasible generalized least squares (FGLS) version of the two-stage least squares
estimation leads to consistent and asymptotically more efficient estimates.
This estimation procedure is generally called °three-stage least
squares° \citep[3SLS,][]{zellner62b}.
The standard 3SLS estimator and its covariance matrix are obtained by
Equations~\ref{eq:2sls-w2sls-3sls} and~\ref{eq:cov-2sls-w2sls-3sls}
with $\OHat$ equal to $\SHat \otimes I_T$,
where $\SHat$ is the estimated covariance matrix of the disturbance terms.

@@

A further approach to obtain a residual covariance
matrix is
\begin{eqnarray}
   \sHat_{ij} & = &
   \frac{ \uHat_i^\top \uHat_j }
   { T - K_i - K_j + \tr \left[ X_i \left( X_i^\top X_i \right)^{-1}
   X_i^\top X_j \left( X_j^\top X_j \right)^{-1} X_j^\top \right] }
   \label{eq:rcovTheil} \\
   & = &
   \frac{ \uHat_i^\top \uHat_j }
   { T - K_i - K_j + \tr \left[ \left( X_i^\top X_i \right)^{-1}
   X_i^\top X_j \left( X_j^\top X_j \right)^{-1} X_j^\top X_i \right] }
\end{eqnarray} 
\citep[p.~309]{zellner62c}.
This yields an unbiased estimator for all elements of $\Sigma$,
but even if $\SHat$ is an unbiased estimator of $\Sigma$,
its inverse $\SHat^{-1}$ is not an unbiased estimator
of $\Sigma^{-1}$ \citep[p.~322]{theil71}.
Furthermore, the covariance matrix calculated by Equation~\ref{eq:rcovTheil}
is not necessarily positive semidefinite \citep[p.~322]{theil71}.
Hence, °it is doubtful whether [this formula] is really superior to 
[Equation~\ref{eq:rcovNoDfCor}]° \citep[p.~322]{theil71}.

@@

<<echo=FALSE>>=
options( prompt = "R> " )
@

@@

Finally, the package includes some internal (hidden) helper functions:
\codeD{.prepareData}{systemfit},
\code{.stackMatList}, and \code{.prepareWmatrix}
for preparing the data matrices;
\code{.calcXtOmegaInv} and \code{.calcGLS}
for calculating the GLS estimator;
\code{.calcResidCov} and \code{.calcSigma2}
for calculating the (co)variances of the residuals;
and \codeD{.ftest}{systemfit}
for calculating \citeauthor{theil71}'s $F$ statistic.
If \code{systemfit} is applied to
a (classical) °seemingly unrelated regression° analysis with panel data,
it calls the hidden internal function \code{.systemfitPanel},
which reshapes the data, creates the formulas to be estimated,
and --- if requested --- specifies restrictions
to ensure that the coefficients of all individuals are equal.

@@

Since version 1.0, the \code{systemfit} function by default
uses the \pkg{Matrix} package \citep{r-matrix-07}
for all computations where matrices are involved.
The \pkg{Matrix} package provides classes for different types of matrices.
For instance, we choose class \code{dgeMatrix}
(°real matrices in general storage mode°),
for matrices $X_i$ in Equation~\ref{eq:model-array},
class \code{dgCMatrix} (°general, numeric, sparse matrices
in the (sorted) compressed sparse column format°)
for matrix $X$ in Equation~\ref{eq:model-matrices},
and class \code{dspMatrix}
(°symmetric real matrices in packed storage (one triangle only)°)
for the residual covariance matrix $\SHat$.
If the \pkg{Matrix} package is used,
the possibly huge matrix $\Omega^{-1}$ is no longer a problem,
because it is a sparse matrix
that can be stored in a compressed format (class \code{dgCMatrix}).
Hence, we no longer need the algorithm in Equation~\ref{eq:omegaInv}.
We have tested different ways to calculate a GLS estimator
like in Equation~\ref{eq:ols-wls-sur}
and we found
that the following code is the fastest:
<<eval=FALSE>>=
sigmaInv <- solve( residCov )
xtOmegaInv <- crossprod( xMat, kronecker( sigmaInv, Diagonal( nObs ) ) )
coef <- solve( xtOmegaInv %*% xMat, xtOmegaInv %*%  yVec )
@
In this code snippet,
\code{residCov} is the residual covariance matrix $\SHat$,
\code{nObs} is the number of observations in each equation $T$,
\code{xMat} is the matrix $X$ and
\code{yVec} is the vector $y$ in Equation~\ref{eq:ols-wls-sur}.

@@

Function \code{sem} (package \pkg{sem}, \citealp{r-sem-0.9-8})
can be used to estimate structural equation models in \proglang{R}
by limited information maximum likelihood (LIML)
and full information maximum likelihood (FIML)
assuming normal or multinormal errors, respectively.
A special feature of this function is the estimation
of models with unobserved (°latent°) variables,
which is not possible with \code{systemfit}.
While \code{sem} cannot be used to consistently estimate
systems of simultaneous equations with some endogenous regressors,
it can be used to estimate systems of equations,
where all regressors are exogenous.
However, the latter is rather cumbersome
(see appendix~\ref{sec:sem}).
Hence, \code{systemfit} is the only function in \proglang{R}
that can be used to estimate systems of simultaneous equations
and it is the most convenient function to estimate
systems of equations with purely exogenous regressors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Using systemfit]{Using \pkg{systemfit}}\label{sec:Usage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
options( prompt = "R> " )
@

In this section we demonstrate how to use the \pkg{systemfit} package.
First, we show the standard usage of \code{systemfit} by a simple example.
Second, several options that can be specified by the user are presented.
Then, the usage of \code{systemfit}
for a (classical) °seemingly unrelated regression° analysis
with panel data is described.
Finally, we demonstrate how to apply some statistical tests.

@@

The optional argument \code{method} is a string
that determines the estimation method.
It must be either \code{"OLS"}, \code{"WLS"}, \code{"SUR"},
\code{"2SLS"}, \code{"W2SLS"}, or \code{"3SLS"}.
These methods correspond to the estimation methods
described in Sections~\ref{sec:Estimation-ols-wls-sur},
\ref{sec:Estimation-2sls-w2sls-3sls}, and~\ref{sec:Restrictions}.
The following command estimates the model described above as
°seemingly unrelated regression°.
<<systemfit_usage>>=
fitsur <- systemfit( eqSystem, method = "SUR" )
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Instrumental variables}
The instruments for a 2SLS, W2SLS or 3SLS estimation can be specified
by the argument \code{inst}.
If the same instruments should be used for all equations,
\code{inst} must be a one-sided formula.%
\footnote{
A one-sided formula is a standard formula in \proglang{R}
without a dependent variable.}
If different instruments should be used for each equation,
\code{inst} must be a list that contains a one-sided formula
for each equation.
The following example uses instrumental variables
to estimate the model described above
by °three-stage least squares° (3SLS).
While the first command specifies the same instruments for all equations,
the second uses different instruments:
<<systemfit_usage>>=
fit3sls  <- systemfit( eqSystem, method = "3SLS",
   inst = ~ income + farmPrice + trend )
fit3sls2 <- systemfit( eqSystem, method = "3SLS",
   inst = list( ~ farmPrice + trend, ~ income + farmPrice + trend ) )
@

@@

The \code{systemfit} function can also be used
for a (classical) °seemingly unrelated regression° analysis
with panel data.
For this type of analysis,
the data must be provided in a transformed data frame of class \codeD{plm}{dim}%
\footnote{
Generally, panel data can be either in °long format°
(different individuals are arranged below each other)
or in °wide format°
(different individuals are arranged next to each other).
For this analysis, the data must be in °long format°.
}
which can be created with the function \codeD{plm}{data}
from the \proglang{R} package \pkg{plm}
\citep{r-plm-0.3-1}.
In contrast to the previously described usage of \code{systemfit},
argument \code{formula} must be a single equation (object of class \code{formula}).
This formula is estimated for all individuals.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Replication of textbook results}\label{sec:reliability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
options( prompt = "R> " )
@

@@

Second, we try to replicate Klein's °Model I° \citep{klein50}
that is described in \citet[p.~381]{greene03}.
The data are available from the online complements to \citet{greene03},
Table~F15.1
(\url{http://pages.stern.nyu.edu/~wgreene/Text/tables/TableF15-1.txt}),
and the estimation results are presented in Table~15.3 (p.~412).

@@

Third, we reproduce Example~14.1 of \citet[p.~340]{greene03}
that is based on \citet{grunfeld58}.
The data are available from the online complements to \citet{greene03},
Table~F13.1
(\url{http://pages.stern.nyu.edu/~wgreene/Text/tables/TableF13-1.txt}).
Several different versions of the °Grunfeld° data set can be found,
whereas the version of \citet{greene03}
is considered incorrect \citep{cummins01}.
However, we use this incorrect version to replicate the results
in \citet{greene03}, Tables~14.1 and~14.2 (p.~351).%
\footnote{
A correct version of this data set with five additional firms
is available as data set \code{Grunfeld} in the \pkg{Ecdat} package
\citep{r-Ecdat-0.1-5}.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Computation times}\label{sec:timings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
options( prompt = "R> " )
@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Estimating systems of equations with sem]
   {Estimating systems of equations with \code{sem}}
\label{sec:sem}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
options( prompt = "R> " )
options( width = 75 )
@

This section compares the commands to estimate a system of equations
by \code{sem} and \code{systemfit}.
This comparison uses Klein's °Model I° (see Section~\ref{sec:KleinsModel}).
Before starting the estimation,
we load the \pkg{sem} and \pkg{systemfit} package
as well as the required data set.
<<systemfit_sem>>=
library( "sem" )
library( "systemfit" )
data( "KleinI" )
@

>>>>

\documentclass[article,nojss]{jss}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{float}

@@

% initialisation stuff
\SweaveOpts{engine=R}
%\VignetteIndexEntry{Systemfit}
%\VignetteDepends{plm,sem}
%\VignetteKeywords{R, system of simultaneous equations,
% seemingly unrelated regression, two-stage least squares,
% three-stage least squares, instrumental variables}
%\VignettePackage{systemfit}
<<echo=FALSE>>=
options( prompt = "R> ", ctinue = "+  " )
@

@@

Many theoretical models that are econometrically estimated
consist of more than one equation.
The disturbance terms of these equations are likely
to be contemporaneously correlated,
because unconsidered factors
that influence the disturbance term in one equation
probably influence the disturbance terms in other equations, too.
Ignoring this contemporaneous correlation
and estimating these equations separately
leads to inefficient estimates of the coefficients.
However, estimating all equations simultaneously
with a ``generalized least squares'' (GLS) estimator,
which takes the covariance structure of the residuals into account,
leads to efficient estimates.
This estimation procedure is generally called
``seemingly unrelated regression'' \citep[SUR,][]{zellner62}.
Another reason to estimate a system of equations simultaneously are
cross-equation restrictions on the coefficients.%
\footnote{
Especially the economic theory suggests many cross-equation
restrictions on the coefficients (e.g., the symmetry restriction in demand models).
}
Estimating the coefficients under cross-equation restrictions
and testing these restrictions
requires a simultaneous estimation approach.

Furthermore, these models can contain variables
that appear on the left-hand side in one equation
and on the right-hand side of another equation.
Ignoring the endogeneity of these variables can lead to inconsistent
estimates.
This simultaneity bias can be corrected for by
applying a ``two-stage least squares'' (2SLS) estimation
to each equation.
Combining this estimation method with the SUR method results
in a simultaneous estimation of the system of equations
by the ``three-stage least squares'' (3SLS) method
\citep{zellner62b}.

@@

The \pkg{systemfit} package provides the capability to estimate
systems of linear equations in \proglang{R}
\citep{r-project-07}.
Currently, the estimation methods
``ordinary least squares'' (OLS),
``weighted least squares'' (WLS),
``seemingly unrelated regression'' (SUR),
``two-stage least squares'' (2SLS),
``weighted two-stage least squares'' (W2SLS), and
``three-stage least squares'' (3SLS)
are implemented.%
\footnote{
In this context,
the term ``weighted'' in ``weighted least squares'' (WLS) and
``weighted two-stage least squares'' (W2SLS) means
that the \emph{equations} might have different weights
and \emph{not}
that the \emph{observations} have different weights.
}
The WLS, SUR, W2SLS, and 3SLS estimates can be based
either on one-step (OLS or 2SLS) (co)variances
or these estimations can be iterated,
where the (co)variances are calculated from the estimates of the previous step.
Furthermore,
the \pkg{systemfit} package
provides statistical tests for restrictions on the coefficients
and for testing the consistency of the 3SLS estimation.

@@

We can write the ``stacked'' system as
\begin{equation}
   \left[ \begin{array}{c}
      y_1 \\ y_2\\ \vdots\\ y_G
   \end{array} \right] = 
   \left[ \begin{array}{cccc}
      X_1 & 0 & \cdots & 0\\
      0 & X_2 & \cdots & 0\\
      \vdots & \vdots & \ddots & \vdots\\
      0 & 0 & \cdots & X_G
   \end{array}\right]
   \left[ \begin{array}{c}
      \beta_1 \\ \beta_2 \\ \vdots\\ \beta_G
   \end{array} \right] +
   \left[ \begin{array}{c}
      u_1 \\ u_2 \\ \vdots\\ u_G
   \end{array} \right]
   \label{eq:model-array}
\end{equation}

@@

If the disturbances are contemporaneously correlated, a generalized
least squares (GLS) estimation leads to an efficient estimator for the coefficients.
In this case, the GLS estimator is generally called ``seemingly unrelated regression''
(SUR) estimator \citep{zellner62}.
However, the true covariance matrix of the disturbance terms
is generally unknown.
The textbook solution for this problem is
a feasible generalized least squares (FGLS) estimation.
As the FGLS estimator is based on
an estimated covariance matrix of the disturbance terms,
it is only asymptotically efficient.
In case of a SUR estimator, $\OHat$ in Equations~\ref{eq:ols-wls-sur}
and~\ref{eq:cov-ols-wls-sur}
is $\SHat \otimes I_T$,
where $\SHat$ is the estimated covariance matrix of the disturbance terms.

@@

At the first stage new (``fitted'') regressors are obtained by
\begin{equation}
   \XHat_i = Z_i \left( Z_i^\top Z_i \right)^{-1} Z_i^\top X_i .
\end{equation}
Then, these ``fitted'' regressors are substituted for the original regressors
in Equation~\ref{eq:ols-wls-sur}
to obtain unbiased 2SLS, W2SLS, or 3SLS estimates of $\beta$ by
\begin{equation}
   \bHat
   = \left( \XHat^\top \OHat^{-1} \XHat \right)^{-1} \XHat^\top \OHat^{-1} y ,
   \label{eq:2sls-w2sls-3sls}
\end{equation}
where
\begin{equation}
   \XHat =
   \left[ \begin{array}{cccc}
      \XHat_1 & 0       & \cdots & 0\\
      0       & \XHat_2 & \cdots & 0\\
      \vdots  & \vdots  & \ddots & \vdots\\
      0       & 0       & \cdots & \XHat_G
   \end{array}\right] .
\end{equation}
An estimator of the covariance matrix of the estimated coefficients
can be obtained from Equation~\ref{eq:cov-ols-wls-sur} analogously.
Hence, we get
\begin{equation}
   \COVHat \left[ \bHat \right]
   = \left( \XHat^\top \OHat^{-1} \XHat \right)^{-1} .
   \label{eq:cov-2sls-w2sls-3sls}
\end{equation}

@@

If the disturbances are contemporaneously correlated,
a feasible generalized least squares (FGLS) version of the two-stage least squares
estimation leads to consistent and asymptotically more efficient estimates.
This estimation procedure is generally called ``three-stage least
squares'' \citep[3SLS,][]{zellner62b}.
The standard 3SLS estimator and its covariance matrix are obtained by
Equations~\ref{eq:2sls-w2sls-3sls} and~\ref{eq:cov-2sls-w2sls-3sls}
with $\OHat$ equal to $\SHat \otimes I_T$,
where $\SHat$ is the estimated covariance matrix of the disturbance terms.

@@

A further approach to obtain a residual covariance
matrix is
\begin{eqnarray}
   \sHat_{ij} & = &
   \frac{ \uHat_i^\top \uHat_j }
   { T - K_i - K_j + \tr \left[ X_i \left( X_i^\top X_i \right)^{-1}
   X_i^\top X_j \left( X_j^\top X_j \right)^{-1} X_j^\top \right] }
   \label{eq:rcovTheil} \\
   & = &
   \frac{ \uHat_i^\top \uHat_j }
   { T - K_i - K_j + \tr \left[ \left( X_i^\top X_i \right)^{-1}
   X_i^\top X_j \left( X_j^\top X_j \right)^{-1} X_j^\top X_i \right] }
\end{eqnarray} 
\citep[p.~309]{zellner62c}.
This yields an unbiased estimator for all elements of $\Sigma$,
but even if $\SHat$ is an unbiased estimator of $\Sigma$,
its inverse $\SHat^{-1}$ is not an unbiased estimator
of $\Sigma^{-1}$ \citep[p.~322]{theil71}.
Furthermore, the covariance matrix calculated by Equation~\ref{eq:rcovTheil}
is not necessarily positive semidefinite \citep[p.~322]{theil71}.
Hence, ``it is doubtful whether [this formula] is really superior to
[Equation~\ref{eq:rcovNoDfCor}]'' \citep[p.~322]{theil71}.

@@

@@

Finally, the package includes some internal (hidden) helper functions:
\codeD{.prepareData}{systemfit},
\code{.stackMatList}, and \code{.prepareWmatrix}
for preparing the data matrices;
\code{.calcXtOmegaInv} and \code{.calcGLS}
for calculating the GLS estimator;
\code{.calcResidCov} and \code{.calcSigma2}
for calculating the (co)variances of the residuals;
and \codeD{.ftest}{systemfit}
for calculating \citeauthor{theil71}'s $F$ statistic.
If \code{systemfit} is applied to
a (classical) ``seemingly unrelated regression'' analysis with panel data,
it calls the hidden internal function \code{.systemfitPanel},
which reshapes the data, creates the formulas to be estimated,
and --- if requested --- specifies restrictions
to ensure that the coefficients of all individuals are equal.

@@

Since version 1.0, the \code{systemfit} function by default
uses the \pkg{Matrix} package \citep{r-matrix-07}
for all computations where matrices are involved.
The \pkg{Matrix} package provides classes for different types of matrices.
For instance, we choose class \code{dgeMatrix}
(``real matrices in general storage mode''),
for matrices $X_i$ in Equation~\ref{eq:model-array},
class \code{dgCMatrix} (``general, numeric, sparse matrices
in the (sorted) compressed sparse column format'')
for matrix $X$ in Equation~\ref{eq:model-matrices},
and class \code{dspMatrix}
(``symmetric real matrices in packed storage (one triangle only)'')
for the residual covariance matrix $\SHat$.
If the \pkg{Matrix} package is used,
the possibly huge matrix $\Omega^{-1}$ is no longer a problem,
because it is a sparse matrix
that can be stored in a compressed format (class \code{dgCMatrix}).
Hence, we no longer need the algorithm in Equation~\ref{eq:omegaInv}.
We have tested different ways to calculate a GLS estimator
like in Equation~\ref{eq:ols-wls-sur}
and we found
that the following code is the fastest:
<<eval=FALSE>>=
sigmaInv <- solve( residCov )
xtOmegaInv <- crossprod( xMat, kronecker( sigmaInv, Diagonal( nObs ) ) )
coef <- solve( xtOmegaInv %*% xMat, xtOmegaInv %*%  yVec )
@
In this code snippet,
\code{residCov} is the residual covariance matrix $\SHat$,
\code{nObs} is the number of observations in each equation $T$,
\code{xMat} is the matrix $X$ and
\code{yVec} is the vector $y$ in Equation~\ref{eq:ols-wls-sur}.

@@

Function \code{sem} (package \pkg{sem}, \citealp{r-sem-0.9-8})
can be used to estimate structural equation models in \proglang{R}
by limited information maximum likelihood (LIML)
and full information maximum likelihood (FIML)
assuming normal or multinormal errors, respectively.
A special feature of this function is the estimation
of models with unobserved (``latent'') variables,
which is not possible with \code{systemfit}.
While \code{sem} cannot be used to consistently estimate
systems of simultaneous equations with some endogenous regressors,
it can be used to estimate systems of equations,
where all regressors are exogenous.
However, the latter is rather cumbersome
(see appendix~\ref{sec:sem}).
Hence, \code{systemfit} is the only function in \proglang{R}
that can be used to estimate systems of simultaneous equations
and it is the most convenient function to estimate
systems of equations with purely exogenous regressors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Using systemfit]{Using \pkg{systemfit}}\label{sec:Usage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section we demonstrate how to use the \pkg{systemfit} package.
First, we show the standard usage of \code{systemfit} by a simple example.
Second, several options that can be specified by the user are presented.
Then, the usage of \code{systemfit}
for a (classical) ``seemingly unrelated regression'' analysis
with panel data is described.
Finally, we demonstrate how to apply some statistical tests.

@@

The optional argument \code{method} is a string
that determines the estimation method.
It must be either \code{"OLS"}, \code{"WLS"}, \code{"SUR"},
\code{"2SLS"}, \code{"W2SLS"}, or \code{"3SLS"}.
These methods correspond to the estimation methods
described in Sections~\ref{sec:Estimation-ols-wls-sur},
\ref{sec:Estimation-2sls-w2sls-3sls}, and~\ref{sec:Restrictions}.
The following command estimates the model described above as
``seemingly unrelated regression''.
<<systemfit_usage>>=
fitsur <- systemfit( eqSystem, method = "SUR" )
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Instrumental variables}
The instruments for a 2SLS, W2SLS or 3SLS estimation can be specified
by the argument \code{inst}.
If the same instruments should be used for all equations,
\code{inst} must be a one-sided formula.%
\footnote{
A one-sided formula is a standard formula in \proglang{R}
without a dependent variable.}
If different instruments should be used for each equation,
\code{inst} must be a list that contains a one-sided formula
for each equation.
The following example uses instrumental variables
to estimate the model described above
by ``three-stage least squares'' (3SLS).
While the first command specifies the same instruments for all equations,
the second uses different instruments:
<<systemfit_usage>>=
fit3sls  <- systemfit( eqSystem, method = "3SLS",
   inst = ~ income + farmPrice + trend )
fit3sls2 <- systemfit( eqSystem, method = "3SLS",
   inst = list( ~ farmPrice + trend, ~ income + farmPrice + trend ) )
@

@@

The \code{systemfit} function can also be used
for a (classical) ``seemingly unrelated regression'' analysis
with panel data.
For this type of analysis,
the data must be provided in a transformed data frame of class \codeD{plm}{dim}%
\footnote{
Generally, panel data can be either in ``long format''
(different individuals are arranged below each other)
or in ``wide format''
(different individuals are arranged next to each other).
For this analysis, the data must be in ``long format''.
}
which can be created with the function \codeD{plm}{data}
from the \proglang{R} package \pkg{plm}
\citep{r-plm-0.3-1}.
In contrast to the previously described usage of \code{systemfit},
argument \code{formula} must be a single equation (object of class \code{formula}).
This formula is estimated for all individuals.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Replication of textbook results}\label{sec:reliability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Second, we try to replicate Klein's ``Model I'' \citep{klein50}
that is described in \citet[p.~381]{greene03}.
The data are available from the online complements to \citet{greene03},
Table~F15.1
(\url{http://pages.stern.nyu.edu/~wgreene/Text/tables/TableF15-1.txt}),
and the estimation results are presented in Table~15.3 (p.~412).

@@

Third, we reproduce Example~14.1 of \citet[p.~340]{greene03}
that is based on \citet{grunfeld58}.
The data are available from the online complements to \citet{greene03},
Table~F13.1
(\url{http://pages.stern.nyu.edu/~wgreene/Text/tables/TableF13-1.txt}).
Several different versions of the ``Grunfeld'' data set can be found,
whereas the version of \citet{greene03}
is considered incorrect \citep{cummins01}.
However, we use this incorrect version to replicate the results
in \citet{greene03}, Tables~14.1 and~14.2 (p.~351).%
\footnote{
A correct version of this data set with five additional firms
is available as data set \code{Grunfeld} in the \pkg{Ecdat} package
\citep{r-Ecdat-0.1-5}.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Computation times}\label{sec:timings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Estimating systems of equations with sem]
   {Estimating systems of equations with \code{sem}}
\label{sec:sem}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
options( width = 75 )
@

This section compares the commands to estimate a system of equations
by \code{sem} and \code{systemfit}.
This comparison uses Klein's ``Model I'' (see Section~\ref{sec:KleinsModel}).
Before starting the estimation,
we load the \pkg{sem} and \pkg{systemfit} package
as well as the required data set.
<<systemfit_sem>>=
library( "sem" )
library( "systemfit" )
data( "KleinI" )
@
